# Immune deconvolution of TCGA Pan-Cancer from the XenaBrowser

## Load libraries:

```{r}
library(tidyverse)
library(R.utils)
library(immunedeconv)
library(stringr)
library(clusterProfiler)
library(org.Hs.eg.db)
library(cowplot)
#library(biomaRt)
#library(BiocFileCache)
#library(BiocManager)
```

## Download dataset:

```{r}
cohort = "TCGA_PANCAN"
file_path <- "Data/XenaBrowser/"
dir.create(file_path, showWarnings = FALSE, recursive = TRUE)
data_type = "gene_expression_RNAseq_tpm"

## Specify the URL of the file you want to download
url <- "https://toil-xena-hub.s3.us-east-1.amazonaws.com/download/tcga_RSEM_gene_tpm.gz"

# Specify the file name and location where you want to save the file on your computer
file_name_data <- paste0(cohort, "_", data_type)

options(timeout = 10000)

if (!(file.exists(paste(file_path, file_name_data, ".tsv.gz", sep = "")))){
  download.file(url, paste(file_path, file_name_data, ".tsv.gz", sep = ""), mode = "wb")
}

if (!(file.exists(paste(file_path, file_name_data, ".tsv", sep = "")))){
  gunzip(paste(file_path, file_name_data, ".tsv.gz", sep = ""))
}
```

## Import data:

```{r}
df_data <- as_tibble(read.csv(paste(file_path, file_name_data, ".tsv", sep = ""),sep = "\t")) 
df_phenotype_disease <- as_tibble(read.csv(paste(file_path, file_name_data, ".tsv", sep = ""),sep = "\t"))
```

## Data wrangling:

Extract sample IDs for specific cancer types:

```{r}
## Get sample IDs for AML:
AML_samples <- as_tibble(df_phenotype_disease) |>
  filter(X_primary_disease == "acute myeloid leukemia") |>
  dplyr::select(sample)

SCM_samples <- as_tibble(df_phenotype_disease) |>
  filter(X_primary_disease == "skin cutaneous melanoma") |>
  dplyr::select(sample)

PAD_samples <- as_tibble(df_phenotype_disease) |>
  filter(X_primary_disease == "pancreatic adenocarcinoma") |>
  dplyr::select(sample)

disease_list <- df_phenotype_disease |>
  distinct(X_primary_disease)
```

Filter the expression data frame to include only samples from selected cancer type:

```{r}
## Get only samples from AML cancer:
df_data_gene_expr_AML <- df_data |>
  rename_with(~ str_replace_all(., "\\.", "-")) |>
  dplyr::select(sample, any_of(AML_samples$sample)) |>
  dplyr::rename(gene = sample) |>
  mutate(gene = str_replace(gene, "\\..*", ""))

df_data_gene_expr_SCM <- df_data |>
  rename_with(~ str_replace_all(., "\\.", "-")) |>
  dplyr::select(sample, any_of(SCM_samples$sample)) |>
  dplyr::rename(gene = sample) |>
  mutate(gene = str_replace(gene, "\\..*", ""))

df_data_gene_expr_PAD <- df_data |>
  rename_with(~ str_replace_all(., "\\.", "-")) |>
  dplyr::select(sample, any_of(PAD_samples$sample)) |>
  dplyr::rename(gene = sample) |>
  mutate(gene = str_replace(gene, "\\..*", ""))
```

```{r}
target_list <- c("CD274", "PDCD1LG2", "IDO1", "TDO2", "ARG1", "ARG2", "TGFB1", "CCL22", "CCL2", "LGALS3", "LRRC32", "IL10", "SIGLEC15", "BIRC7")
```

Convert Ensemble IDs to HGNC symbols:

```{r}
# Define your list of Ensembl gene IDs
#ensembl_genes <- df_data_gene_expr$gene
#gene_mapping <- bitr(ensembl_genes, fromType = "ENSEMBL", toType = "SYMBOL", OrgDb = "org.Hs.eg.db")
```

```{r}
gene_expr_df_AML <- as_tibble(gene_mapping) |>
  dplyr::rename(gene = ENSEMBL, symbol = SYMBOL) |>
  inner_join(df_data_gene_expr_AML, by="gene") |>
  dplyr::select(symbol, everything(), -gene) |>
  distinct(symbol, .keep_all = TRUE) |>
  #mutate(across(-1, as.numeric)) |>
  #mutate(across(-1, ~ (2^.x) - 0.001)) |>
  column_to_rownames(var = "symbol")

gene_expr_df_SCM <- as_tibble(gene_mapping) |>
  dplyr::rename(gene = ENSEMBL, symbol = SYMBOL) |>
  inner_join(df_data_gene_expr_SCM, by="gene") |>
  dplyr::select(symbol, everything(), -gene) |>
  distinct(symbol, .keep_all = TRUE) |>
  #mutate(across(-1, as.numeric)) |>
  #mutate(across(-1, ~ (2^.x) - 0.001)) |>
  column_to_rownames(var = "symbol")

gene_expr_df_PAD <- as_tibble(gene_mapping) |>
  dplyr::rename(gene = ENSEMBL, symbol = SYMBOL) |>
  inner_join(df_data_gene_expr_PAD, by="gene") |>
  dplyr::select(symbol, everything(), -gene) |>
  distinct(symbol, .keep_all = TRUE) |>
  #mutate(across(-1, as.numeric)) |>
  #mutate(across(-1, ~ (2^.x) - 0.001)) |>
  column_to_rownames(var = "symbol")
```

Convert the gene expression data frame as a matrix:

```{r}
gene_expr_matrix_AML <- as.matrix(gene_expr_df_AML)
gene_expr_matrix_SCM <- as.matrix(gene_expr_df_SCM)
gene_expr_matrix_PAD <- as.matrix(gene_expr_df_PAD)
```

## Deconvolution:

Using ESTIMATE which computes a score for the tumoral, immune and stromal components and the fraction of tumor purity.

<https://bioinformatics.mdanderson.org/estimate/>

```{r}
estimate_AML = deconvolute_estimate(gene_expr_matrix_AML)
estimate_SCM = deconvolute_estimate(gene_expr_matrix_SCM)
estimate_PAD = deconvolute_estimate(gene_expr_matrix_PAD)
```

Plot estimates as a boxplot:

```{r}
estimate_PAD |>
  rownames_to_column(var = "estimates") |>
  pivot_longer(cols = -estimates,
               names_to = "sample",
               values_to = "scores") |>
  ggplot(aes(x = scores, fill = estimates)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  facet_wrap(~ estimates, scales = "free_x") +  # Separate histograms by estimates
  theme_minimal() +
  labs(x = "Scores", y = "Count", title = "Distribution of Scores by ESTIMATE in Pancreatic Adenocarcinoma")
```

Using quanTIseq to estimate immune cell fractions:

<https://icbi.i-med.ac.at/software/quantiseq/doc/index.html>

```{r}
quantiseq_AML <- deconvolute(gene_expr_matrix_AML, "quantiseq", tumor = TRUE)
quantiseq_SCM <- deconvolute(gene_expr_matrix_SCM, "quantiseq", tumor = TRUE)
quantiseq_PAD <- deconvolute(gene_expr_matrix_PAD, "quantiseq", tumor = TRUE)
```

Plotting the cell fractions in a stacked barchart:

```{r}
quantiseq_AML %>%
  gather(sample, fraction, -cell_type) %>%
  # plot as stacked bar chart
  ggplot(aes(x = sample, y = fraction, fill = cell_type)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_brewer(palette = "Paired") +
  theme(axis.text.y = element_blank())
```

```{r}
quantiseq_PAD %>%
  gather(sample, fraction, -cell_type) %>%
  ggplot(aes(x=cell_type, y=fraction)) + 
  geom_boxplot() +
  labs(
    title = "Pancreatic Adenocarcinoma",
    subtitle = "Distribution of cell populations",
    caption = "Source: Xena Browser, TCGA PANCAN",
    x = "Cell types",
    y = "Fraction"
  ) +
  theme(
    plot.title = element_text(color = "#0099f8", size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(face = "bold.italic", hjust = 0.5),
    plot.caption = element_text(face = "italic"),
    axis.text.x = element_text(angle = 45, hjust = 1))
```

MCP-counter provides scores in arbitrary units that are only comparable between samples, but not between cell-types. With the scores being in arbitrary units, the results are not useful for judging if a cell type is present in the sample, or not. However, we can compare the relative values between samples and relate them to the results we obtained earlier using quanTIseq.

```{r}
mcp_counter_AML <- deconvolute(gene_expr_df_AML, "mcp_counter")
mcp_counter_SCM <- deconvolute(gene_expr_df_SCM, "mcp_counter")
mcp_counter_PAD <- deconvolute(gene_expr_df_PAD, "mcp_counter")
```

```{r, fig.height=5, fig.width=12, warning=FALSE}
mcp_counter_SCM %>%
  gather(sample, score, -cell_type) %>%
  ggplot(aes(x = sample, y = score, color = cell_type)) +
  geom_point(size = 1) +
  facet_wrap(~cell_type, scales = "free_x", ncol = 6) +
  scale_color_brewer(palette = "Paired", guide = FALSE) +
  coord_flip() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        axis.text.y = element_blank()) + 
  labs(title = "Skin Cutaneous Melanoma")
```

Predict cell population-specific gene expression 

```{r}
# Define a function to perform regression and predict gene expression for all genes
predict_all_genes <- function(gene_expr_df, deconvolution) {
  # Ensure deconvolution is a data frame and extract relevant data
  deconvolution <- as.data.frame(deconvolution)
  cell_types <- deconvolution$cells
  deconvolution_samples <- deconvolution[, -1]  # Remove the cell types column

  # Create an empty matrix to store predictions
  predicted_matrix <- matrix(nrow = nrow(gene_expr_df), ncol = ncol(gene_expr_df))
  rownames(predicted_matrix) <- rownames(gene_expr_df)
  colnames(predicted_matrix) <- colnames(gene_expr_df)
  
  # Iterate over each gene
  for (gene in rownames(gene_expr_df)) {
    # Extract gene expression values for the gene of interest
    gene_expression <- as.numeric(gene_expr_df[gene, ])
    
    # Check if the length of gene_expression matches the number of columns in deconvolution_samples
    if (length(gene_expression) == ncol(deconvolution_samples)) {
      # Perform linear regression
      model <- lm(gene_expression ~ t(deconvolution_samples))
      
      # Predict gene expression
      predicted_expression <- predict(model, newdata = as.data.frame(t(deconvolution_samples)))
      
      # Store predictions in the matrix
      predicted_matrix[gene, ] <- predicted_expression
    } else {
      warning(paste("Length of gene expression vector does not match number of columns in Deconvolution for gene:", gene))
    }
  }
  
  return(predicted_matrix)
}
```

Use the function to make predicted gene expression matrix:

```{r}
# Making a subset of the gene expression matrix so the correction is only for those genes:
gene_expr_df_AML_targets <- gene_expr_df_AML[rownames(gene_expr_df_AML) %in% target_list, ]
gene_expr_df_PAD_targets <- gene_expr_df_PAD[rownames(gene_expr_df_PAD) %in% target_list, ]
gene_expr_df_SCM_targets <- gene_expr_df_SCM[rownames(gene_expr_df_SCM) %in% target_list, ]

predicted_gene_expr_AML <- predict_all_genes(gene_expr_df_AML_targets, quantiseq_AML)
predicted_gene_expr_PAD <- predict_all_genes(gene_expr_df_PAD_targets, quantiseq_PAD)
predicted_gene_expr_SCM <- predict_all_genes(gene_expr_df_SCM_targets, quantiseq_SCM)
```

Plot the gene expression of all targets in the AML data set and compare predicted to original:

```{r}
plt_predicted <- as_tibble(predicted_gene_expr_SCM) |>
  mutate(gene = rownames(predicted_gene_expr_SCM)) |>
  dplyr::select(gene, everything()) |>
  #filter(gene %in% target_list) |>
  #mutate(across(-1, ~ log2(. + 0.001))) |>
  gather(sample, gene_expr, -gene) |>
  ggplot(aes(x=gene_expr, y=gene)) +
  geom_boxplot() +
  labs(title = "Adjusted gene expression SCM",
       x = "log2(tpm+0.001)",
       y = "genes") + 
  geom_vline(xintercept = 0, linetype = "dotted", color = "red") +
  theme(
    plot.title = element_text(color = "#0099f8", size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(face = "bold.italic", hjust = 0.5),
    plot.caption = element_text(face = "italic"))

plt_original <- as_tibble(gene_expr_df_SCM_targets) |>
  mutate(gene = rownames(gene_expr_df_SCM_targets)) |>
  dplyr::select(gene, everything()) |>
  #filter(gene %in% target_list) |>
  #mutate(across(-1, ~ log2(. + 0.001))) |>
  gather(sample, gene_expr, -gene) |>
  ggplot(aes(x=gene_expr, y=gene)) +
  geom_boxplot() +
  labs(title = "Original gene expression SCM",
       x = "log2(tpm+0.001)",
       y = "genes") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "red") +
  theme(
    plot.title = element_text(color = "#0099f8", size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(face = "bold.italic", hjust = 0.5),
    plot.caption = element_text(face = "italic"))
```

```{r}
plot_grid(plt_predicted, plt_original, labels = c('A', 'B'), label_size = 12)
```

```{r}
df_data_long |>
  filter(gene %in% target_list_ENSG) |>
  inner_join(df_phenotype_disease, by=c("sample"))
```

```{r}
estimate <- immunedeconv::deconvolute_estimate(gene_expr_matrix)
```
